---
title: "Lab 03"
author: "ks"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1. Read in the data
First download and then read in with data.table:fread()

```{r}
if (!file.exists("met_all.gz")) {
  download.file("https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz", 
                "met_all.gz", method="libcurl", timeout = 60)  
}

met <- data.table::fread("met_all.gz")
```



## Step 2. Check the dimensions, headers, footers. How many columns, rows are there?


```{r }
dim(met)
head(met)
tail(met)
```
## Step 3. Look at the variables

```{r}
str(met)
```

## Step 4. Take a closer look

```{r}
table(met$day)
```

```{r}
table(met$year)
```

```{r}
table(met$hour)
```

```{r}
summary(met$temp)
summary(met$wind.sp)
```

```{r}
met[,summary(elev)]
```

Replace elevations with 9999 as NA.


```{r}
met[met$elev==9999.0] <- NA
summary(met$elev)
```

The weather station with the highest elevation is at `r max(met$elev,na.rm=TRUE)` meters.

Remove temps below -40 celcius.

```{r}
met <- met[temp>-40][order(temp)]
```

```{r}
met <- met[temp>-15][order(temp)]
```

## Step 5. Check the data against an external data source.

## Step 6. Compute summary statistics

```{r}
met[elev==max(elev,na.rm=TRUE), summary(wind.sp)]
met[elev==max(elev,na.rm=TRUE), summary(temp)]
```

```{r}
met[elev==max(elev,na.rm=TRUE), .(
    temp_wind    = cor(temp,wind.sp,  use = "complete"),
    temp_hour    = cor(temp,hour,     use = "complete"),
    wind.sp_day  = cor(wind.sp,day,    use = "complete"),
    wind.sp_hour = cor(wind.sp,hour,    use = "complete")
)]

```

# Step 7. Exploratory Graphs

Look at key variables

```{r}
hist(met$elev, breaks = 100)
```


```{r}
if (!require(leaflet)) {
  install.packages("leaflet")
  library(leaflet)
}
 
elev <- met[elev==max(elev)]
```

```{r}
leaflet(elev) %>%
  addProviderTiles('OpenStreetMap') %>% 
  addCircles(lat=~lat,lng=~lon, opacity=1, fillOpacity=1, radius=100)
```

